\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}
\vspace{0.5cm}
\hrule
\vspace{0.5cm}

Real-time image and video segmentation is a fundamental preprocessing step in edge computer vision. However, executing multi-dimensional clustering within strict real-time limits ($120+\text{ FPS}$) exceeds the computational capacity of standard CPUs.

This thesis introduces a GPU-accelerated C++/CUDA framework for 5D image segmentation on a joint color-spatial manifold. The system compares a classical Euclidean distance solver against a quantum-inspired engine emulating the quantum SWAP test. GPU bottlenecks are resolved by optimizing thread layouts and memory access patterns within the CUDA kernels. Additionally, an asynchronous Dear ImGui overlay computes performance and quality metrics on a background thread to avoid introducing interface latency.

Tested on an NVIDIA GeForce RTX 5070 Ti, the pipeline runs at up to $750\text{ FPS}$ (achieving at least $259\text{ FPS}$ under the heaviest configuration of $40$ clusters with no downsampling). Due to the monotonic nature of the phase-space mapping, which preserves relative feature ordering, the quantum-inspired backend is mathematically and visually equivalent to classical K-Means, producing identical cluster boundaries and quality metrics under highly contrasting lighting. Finally, profiling shows that compiling the trigonometric calculations into GPU mathematical intrinsics avoids ALU bottlenecks, maintaining speed parity with the classical Euclidean backend.

\vspace{1.5cm}
\noindent \textbf{Keywords:} GPU Acceleration, 5D K-Means, Quantum-Inspired Computing, SWAP Test, Real-Time Video Segmentation.