\chapter{Verification and Testing}
{\itshape A verification pipeline is established to confirm that the classical and quantum-inspired engines are mathematically correct and stable. The automated unit-testing suite is detailed next, followed by the CPU-GPU parity protocol and latency profiling.}

\section{Automated Verification and Robustness Testing}
The core algorithms were verified using GTest \cite{googletest} to test C++ code and CUDA kernels without launching the full GUI application. GTest was chosen because it integrates smoothly with CMake and supports CUDA-based test fixtures. These fixtures automatically reset the GPU and clear device memory before each test, ensuring a clean state and preventing memory leaks from affecting subsequent test results.

The testing suite compiles into a dedicated executable, \texttt{all\_tests}, which mirrors the C++ project structure:
\begin{itemize}
    \item \textbf{Vector Mathematics:} Verifies the accuracy of 5D Euclidean and angular phase distance calculations on both the CPU and GPU.
    \item \textbf{Clustering Engines:} Confirms that the pixel assignment and centroid aggregation logic produce correct results under controlled configurations.
    \item \textbf{Data Preprocessing:} Ensures that raw BGR pixel data is correctly normalized and downsampled according to the stride parameter.
\end{itemize}

All tests execute successfully, achieving a $100\%$ pass rate across the suite.

To guarantee mathematical correctness and stability under extreme operational conditions, unit and robustness tests were implemented to target the GPU kernels and verify host-device memory coordination.

\subsection{Mathematical Correctness and Convergence}
The \texttt{AssignmentCorrectness} and \texttt{QuantumMetricAssignment} tests verify that the engines assign pixels to the correct nearest centroids. This is validated using fixed synthetic datasets with pre-calculated cluster boundaries. 

Iterative stability is also tested through the \texttt{ClassicalConvergence} and \texttt{QuantumConvergence} tests. These use multi-modal data clusters to confirm that both the linear Euclidean and non-linear quantum-inspired engines reach a stable state (where centroids stop moving) within a bounded number of iterations \cite{lloyd1982least}, proving that the phase-based mapping does not cause convergence loops.

\subsection{Boundary Condition Resilience}
A common issue in K-Means is the empty cluster problem, where a centroid is assigned zero pixels during an iteration \cite{aloise2017strategies}, causing a divide-by-zero error during centroid updates. The \texttt{EmptyClusterRobustness} and \texttt{QuantumEmptyClusterRobustness} tests verify that the system detects empty clusters and handles them safely. By placing starting centroids far away from any data points, the tests confirm that the kernels remain stable and do not produce NaN (Not a Number) values, which would otherwise crash the GPU visualization pipeline.

Additionally, the \texttt{HighKAssignment} and \texttt{QuantumHighK} tests evaluate system behavior at the maximum configuration of $K=40$ clusters. This confirms that the shared memory used by the CUDA kernels does not exceed physical hardware bounds, preventing kernel launch failures \cite{CudaGuide}.

\subsection{Cross-Backend Parity and Numerical Drift}
To verify the GPU implementations, a parity protocol is established that compares GPU results against CPU-calculated ground truths using identical data inputs.

During parity testing, small numerical differences (on the order of $10^{-6}$) are expected in the final centroid coordinates due to floating-point drift. This occurs because the CPU and GPU perform float arithmetic differently \cite{CudaGuide, cook2012cuda}. The CPU's floating-point unit often calculates intermediate values in 64-bit or 80-bit precision before rounding back to 32-bit (FP32), while the GPU maintains a strict, high-speed 32-bit pipeline \cite{CudaGuide}. 

Moreover, compiler optimizations like FMA combine multiplications and additions into a single step on the GPU, which alters rounding boundaries compared to sequential CPU steps \cite{CudaGuide}. To handle this hardware-specific behavior, the GTest suite \cite{googletest} uses the \texttt{EXPECT\_NEAR} macro with a tolerance of $10^{-5}$, verifying that the GPU assignments remain mathematically equivalent to the CPU baseline.

\section{Performance Profiling and Experimental Reproducibility}
To measure the pipeline speed and stage latency under real-world conditions, the classical engine was profiled in real-time on a live VGA video feed ($N = 307,200$). By utilizing the system's temporal consistency (warm-starting the current frame's centroids from the previous frame's converged centroids), the optimization iterations are significantly reduced. The resulting sub-millisecond stage latencies and corresponding theoretical frame rates, calculated as the mathematical average of 21 recorded telemetry runs for each configuration, are summarized in Table~\ref{tab:telemetry}.

\begin{table}[htbp]
\centering
\small
\caption[System Performance and Stage Latency Telemetry on RTX 5070 Ti]{System Performance and Latency Telemetry (RTX 5070 Ti GPU)}
\label{tab:telemetry}
\begin{tabular}{ccccccc}
\hline
\textbf{Mode} & \textbf{$K$} & \textbf{$S$} & \textbf{Preprocess} & \textbf{Optimize} & \textbf{Assign} & \textbf{Total Latency} \\ \hline
Performance  & 3  & 8 & 0.28 ms & 0.47 ms & 0.59 ms & \textbf{1.33 ms} \\
Balanced     & 8  & 4 & 0.29 ms & 0.74 ms & 0.59 ms & \textbf{1.62 ms} \\
High Quality & 40 & 1 & 0.30 ms & 2.93 ms & 0.62 ms & \textbf{3.86 ms} \\ \hline
\end{tabular}
\end{table}

The telemetry shows that the pipeline is extremely efficient:
\begin{itemize}
    \item \textbf{Throughput:} Across all three modes (Performance, Balanced, and High Quality), the maximum theoretical throughput is well above the $120\text{ FPS}$ real-time video requirement ($8.33\text{ ms}$ budget) \cite{jerald2009scene}, reaching up to $750\text{ FPS}$, $618\text{ FPS}$, and $259\text{ FPS}$ respectively. This high throughput is directly enabled by the warm-start temporal consistency mechanism, which reduces the required iteration count for convergence to just a few steps per frame on consecutive inputs.
    \item \textbf{Latency scaling:} The preprocessing stage remains highly constant at approximately $0.28\text{ ms}$ to $0.30\text{ ms}$ regardless of the cluster count $K$ and stride $S$, validating the efficiency of the CUDA preprocessor. The optimization (centroid learning) stage scales with the active pixel count (controlled by stride $S$) and cluster count $K$, while the pixel assignment stage scales minimally, from $0.59\text{ ms}$ to $0.62\text{ ms}$, due to CUDA thread block execution efficiency on the GPU.
    \item \textbf{High-density feasibility:} In the heaviest configuration ($K=40, S=1$), which processes all $307,200$ pixels without downsampling, the total execution latency is only $3.86\text{ ms}$ (yielding $259\text{ FPS}$). This confirms that even under maximum computational load on a live video stream, the system operates with substantial processing headroom, leaving the GPU idle for a large portion of the camera's physical frame delivery window of $33.3\text{ ms}$ ($30\text{ FPS}$).
\end{itemize}

\label{sect:scientific_reproducibility}
To ensure that experimental comparisons between the classical and quantum-inspired engines are fair and reproducible, a fixed-seed protocol is implemented. Because K-Means is a stochastic algorithm, varying initial centroids would introduce random noise into convergence and quality benchmarks.

To eliminate this variance, random number generators are configured to use a fixed, global seed value (\texttt{STABLE\_RANDOM\_SEED = 42}). This seed is used to initialize the Mersenne Twister (\texttt{std::mt19937}) engines in the initializers \cite{matsumoto1998mersenne, stroustrup2022tour}, and the Monte Carlo sampling logic in the metrics runner \cite{shutaywi2021silhouette}. This cross-platform consistency guarantees that the starting centroid configurations and subset selections are identical across multiple runs, ensuring comparative benchmarks are scientific and reliable. With these rigorous reproducibility controls established, the next chapter presents the comparative experimental results and detailed quality evaluations of the classical and quantum-inspired engines.
