\chapter{Conclusions and Future Work}\label{chap:conclusions}
{\itshape Real-time, GPU-accelerated video segmentation connects high-level computer vision with performance-constrained execution hardware. This summary details the primary software implementations, empirical results, and future research directions for the classical and quantum-inspired clustering pipelines.}

\section{Summary and Key Findings}
The real-time, GPU-accelerated 5D image and video segmentation pipeline developed in this work integrates classical and quantum-inspired K-Means engines. The primary contributions include:
\begin{enumerate}
    \item \textbf{High-Performance CUDA Pipeline:} An optimized 5D preprocessor and solver in CUDA C++ operating on the BGR-XY manifold \cite{cook2012cuda}. By mapping threads directly to strided grids and caching centroids in low-latency shared memory, the implementation eliminates warp divergence and coalesces VRAM transactions. This design achieves a theoretical throughput of up to $750\text{ FPS}$ on an RTX 5070 Ti, exceeding the $120\text{ FPS}$ ($8.33\text{ ms}$) real-time video requirement \cite{jerald2009scene}.
    \item \textbf{Quantum-Inspired Simulation:} An angular phase-space similarity metric modeled on the multi-qubit state overlap of the quantum SWAP test \cite{buhrman2001quantum}. This backend projects normalized 5D features onto a high-dimensional Hilbert hypersphere via a $[0, \pi/4]$ bounded cosine-squared phase encoding \cite{scholkopf1998nonlinear}.
    \item \textbf{Asynchronous Diagnostics:} An interactive Dear ImGui user interface dashboard \cite{ImGui} and background benchmarking manager. This framework allows on-demand parameter modification (such as cluster count $K$ and stride $S$) and asynchronously calculates execution metrics (latency, iteration counts, WCSS \cite{lloyd1982least}, the Davies-Bouldin ($\mathcal{D}\mathcal{B}$) index \cite{davies1979cluster}, and Monte Carlo-approximated Silhouette scores \cite{rousseeuw1987silhouettes}) on a background thread to prevent display loop stutter.
\end{enumerate}

Empirical evaluations show both engines are viable for real-time edge deployment, with core optimization latency remaining under $3.9\text{ ms}$ ($259\text{ FPS}$) on an NVIDIA RTX 5070 Ti. The quantum-inspired backend achieves complete functional equivalence to classical Euclidean K-Means \cite{lloyd1982least}, producing identical boundaries and quality metrics across all lighting environments. This parity is guaranteed by the monotonic nature of the cosine-squared mapping over the $[0, \pi/4]$ phase range.

Additionally, compiling the angular distance function into the hardware intrinsic \texttt{\_\_cosf()} offloads execution to the GPU's SFUs \cite{cook2012cuda}. This design maintains performance parity with standard ALUs and enables the quantum-inspired engine to run slightly faster under low-density loads.

\section{Future Research Directions}
Future work can extend this framework in three main areas:
\begin{itemize}
    \item \textbf{Spatiotemporal Continuity:} Frame-by-frame centroid drift can be mitigated by incorporating temporal tracking. Subsequent implementations can integrate generalized flow tracking \cite{szeliski2022computer, chauhan2013moving} or topological association \cite{kuhn1955hungarian} to maintain boundary stability across sequential frames while respecting real-time latency constraints.
    \item \textbf{Native Quantum Hardware Execution:} With the maturation of QPUs \cite{preskill2018quantum, bharti2022noisy}, the simulated phase-mapped distance metrics can be ported to physical quantum hardware. This allows for direct evaluation of state-preparation latency, superposition, and entanglement in practical clustering tasks.
    \item \textbf{Edge Device Deployment:} Running high-throughput parallel clustering on resource-constrained embedded systems \cite{slimani2019k} introduces new hardware constraints. Future analysis must quantify the limits of power efficiency, thermal throttling, and memory bandwidth when scaling 5D manifolds on low-wattage architectures.
\end{itemize}